---
title: "Report Air quality and visualisations"
author: "Reneta Todorova up2246775"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Report air quality 

In this report I will look at air pollutants such as PM10, NOx and NO2.
The data that I will analyze and visualize are provided by the British Transport Consortium (BTC) and They have provided data for the Anglesea Road monitoring station which covers the five-year period 2018 to 2023.



```{r echo=FALSE,load_data}

PM10 <- readRDS("PM10.rds")
NO2 <- readRDS("NO2.rds")
NO <- readRDS("NO.rds")
AVG <- readRDS("AVG.rds")
AUGUST <- readRDS("AUGUST.rds")

```



The first pollutant that I will look at is PM10.
PM or particulate matter is not a single pollutant but a collection of many chemical elements and based on (Department for Environment, Food & Rural Affairs, 2024) some of them are toxic and may enter the bloodstream and be transported around the body.
According to (Department for Environment, Food & Rural Affairs, 2024), PM10 refers to particles smaller than 10 micrometres in diameter, making it easy to inhale.


As can be seen from the visualizations below, the concentration of PM 10 varies over the different dates that were requested for analysis. It can be clearly seen that a few days before the first lockdown was announced, the amount of this particulate matter in the air significantly decreased compared to data close to the same date the previous year. An interesting fact is that 3 days after the lockdown was announced, the amounts increased significantly and reached their peak of 48,310 μg/m2. The levels decreased significantly for the next 4 dates that were visualized. As on 29.11.2021 - despite the missing data, we can see again a large spike in the presence of PM10.
From the statements of (Cairn Technology, n.d.) A major source of PM10 includes dust from construction sites and road traffic (particularly diesel cars).
We can only speculate what caused the increased amount of PM10 in the air on March 26, 2020, but perhaps the preparations for the lockdown and the suspension of repair activities caused an increase in the amount of this dust in the air.

```{r , echo=FALSE}
PM10
```



The next element I'm going to look at is NO or Nitric oxide from information of (Department for Environment, Food & Rural Affairs, 2024),
combustion of fuels, particularly in transportation, power generation, and industrial processes.
Looking at the NOx values in 2018 we see high values, which decrease the following year, probably the use of electric cars could reduce these values in the air. 
During the covid pandemic we can see that the values are significantly low, due to the limited travel of people and working from home. 
We observed an interesting spike in levels during the pandemic on the morning of 11/10/2020. When prompted with “what day of the week are the dates: 11/10/2020 and 11/29/2021” and “Are there any significant events in Portsmouth around this date?”, the information generated by ChatGPT can help me analyze the situation ChatGPT (March 2024 version). The UK was under lockdown and the only reason why NOx levels were elevated that day could be due to the intense rush of stores for the upcoming December holidays.
based on my inquiry to the GPT and the answer I received for the date 29.11.2021 - Businesses, events, and social gatherings were fully open with no legal limits. ChatGPT (March 2024 version), which suggests a large flow of people preparing for the upcoming holiday season, after the prolonged stay at home. The levels of this element reach 64.80476. In the following years 2022 and 2023, the levels remain relatively the same, with a slight increase in levels during the morning hours. As these dates are the end of the school year in the UK, we can assume that most families have taken trips during this period.

```{r , echo=FALSE}
NO
```




The last element in the polluted air in this report will be  NO2 or nitrogen dioxide .

NO can react with other gases in the atmosphere to form nitrogen dioxide (NO2) which is harmful to health. (Department for Environment, Food & Rural Affairs, 2024)

With this information and the visualization below, we can clearly notice almost the same relationship between the dates and the increase or decrease in the amount of this element in the air, as with the previously considered element, namely nitric oxide.


```{r , echo=FALSE}
NO2
```



#### Monthly average pollution in 2020
The graph showing Monthly average pollution in 2020 clearly shows that at the beginning of the year the values of all pollutants increased significantly, with most of them reaching maximum values, and then falling before the pandemic. A very interesting spike in the pollutant PM10 was observed in April, assuming that the UK was under COVID lockdown, but if we go back to the table where I looked at the PM10 element individually, we will notice that on 26.03.2020 there was a significant increase. PM 10 is an element that is not a gas but rather dust particles that pollute the air is associated with respiratory conditions as mentioned by (Department for Environment, Food & Rural Affairs, 2024)


```{r , echo=FALSE}
AVG

```



For the last graph, I chose to compare the date August 27, in 2020 at the peak of the COVID pandemic and 2021 after the restrictions were lifted. As in 2021 after the restrictions were lifted, the Victorious Festival was the first major event that gathered many people. As this event attracts people from other cities. From both graphs, you can clearly see the increase in harmful gases, especially in the morning hours and in the evening. As the values exceed double the values during the same period last year.

```{r}
AUGUST
```

 
##                                   **Reference**
 
 Department for Environment, Food & Rural Affairs. (2024, February 15). Emissions of air pollutants in the UK – particulate matter (PM10 and PM2.5). GOV.UK. https://www.gov.uk/government/statistics/emissions-of-air-pollutants/emissions-of-air-pollutants-in-the-uk-particulate-matter-pm10-and-pm25
 
 
 Cairn Technology. (n.d.). What is PM10 air quality? https://cairntechnology.com/what-is-pm10-air-quality/
 
 
 Department for Environment, Food & Rural Affairs. (2024, February 15). Emissions of air pollutants in the UK - nitrogen oxides (NOx). GOV.UK. https://www.gov.uk/government/statistics/emissions-of-air-pollutants/emissions-of-air-pollutants-in-the-uk-nitrogen-oxides-nox
 
 
 OpenAI. (2024). ChatGPT (March 2024 version). [Large language model] https://chatgpt.com/share/67ebc175-bfe0-8006-8b91-e58305d0120c
 
 